---
title: "Workflow: Basics"
number-sections: true
---

## Introduction

This page covers basic concepts when working with R.  I took note for those that were new to me or found useful to remind myself with.


## Clear Workspace

Always start by clearing the workspace. This ensure objects created in other files are not used used here.

```{r}
#| include: true
#| results: 'hide'

rm(list = ls())
```

## Packages {#sec-packages}

### List

List all the packages that will be used in this script.

```{r List of Used Packages}
#| include: true
#| echo: true
#| error: true

packages = c()
```

### Install Missing

Any missing package will be installed automatically. This ensure smoother execution when run by others.

```{r Install Missing Packages}
#| include: true
#| echo: true
#| error: true

# Do NOT modify
install.packages(setdiff(packages, rownames(installed.packages())))
```

### Load

Load all packages

```{r Load Packages}
#| include: true
#| echo: false
#| error: true

# Do NOT modify
lapply(packages, require, character.only = TRUE)
```

## Comments

Use comments to explain the *why* of your code, eg, you changed the default value of a parameter of a function from say `.2` to `.9`, why?


## Nameing Objects Rules

- Allowed characters when naming objects
  - letters
  - numbers
  - `_`
  - `.`
- All names must start with a letter
- R is case-sensitive, ie, `var`, `Var`, and `VAR` are different names
